---
title: "report.Rmd"
author: "Valentina Anthonio"
date: "2023-06-19"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## COVID19 Data Analysis 
This report presents an analysis of COVID19 data, specifically focusing on the number of cases and deaths since the outbreak. The data used for this analysis was sourced from the GitHub account of [John Hopkins](https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/), spanning from 2020 to 2023. Five types of datasets were collected, including global cases, global deaths, US cases, US deaths, and population data which are;

```{r libraries, echo=TRUE}
library(tidyverse)
library(lubridate)
```

### Import files 
```{r load_fikes,echo=TRUE}
url_in <-"https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"
file_names <- c("time_series_covid19_confirmed_US.csv",  "time_series_covid19_confirmed_global.csv", "time_series_covid19_deaths_US.csv",  "time_series_covid19_deaths_global.csv")
urls <-str_c(url_in,file_names)

us_cases <- read_csv(urls[1])
global_cases <-read_csv(urls[2])
us_deaths <- read_csv(urls[3])
global_deaths <- read_csv(urls[4])
```

### Data cleaning
```{r global,echo=TRUE}
#data cleaning and dropping some columns
global_cases <- global_cases %>% pivot_longer(cols = -c('Province/State','Country/Region','Lat','Long'),names_to = 'date',values_to = 'cases') %>% select(-c(Lat,Long))

global_deaths <- global_deaths %>% pivot_longer(cols = -c('Province/State','Country/Region','Lat','Long'),names_to = 'date',values_to = 'deaths') %>% select(-c(Lat,Long))

```

```{r global2, echo=TRUE}
#merge the two global data
global <- global_cases %>% full_join(global_deaths) %>% rename(Country_Region = 'Country/Region', Province_State = 'Province/State') %>% mutate(date=mdy(date))

# filter for cases greater than zero
global <- global %>% filter(cases >0)

global <- global %>% 
  unite("Combined_Key", c(Province_State,Country_Region ), 
        sep =", ",
        na.rm = TRUE, 
        remove=FALSE)

```

```{r US, echo=TRUE}
# perform data cleaning for the US data
us_cases <- us_cases %>% pivot_longer(cols = -(UID:Combined_Key), names_to = "date",values_to = "cases") %>% select(Admin2:cases) %>% mutate(date = mdy(date)) %>% select(-c(Lat,Long_))

us_deaths <- us_deaths %>% pivot_longer(cols = -(UID:Population), names_to = "date",values_to = "deaths") %>% select(Admin2:deaths) %>% mutate(date = mdy(date)) %>% select(-c(Lat,Long_))

#merge the US datasets
US <- us_cases %>% full_join(us_deaths)

```

```{r, global3,echo=TRUE}
#get population data from same site 
uid_lookup_url <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"

uid <-read_csv(uid_lookup_url) %>% 
  select(-c(Lat, Long_,Combined_Key,code3,iso2,iso3,Admin2))

```

```{r, global4, echo=TRUE}
global <- global %>%
  left_join(uid, by= c("Province_State","Country_Region")) %>%
  select(-c(UID,FIPS)) %>%
  select(Province_State,Country_Region,date,cases,deaths,Population,Combined_Key)

```


### Data Analysis
```{r US_totals,echo=TRUE}

US_by_state <- US %>% group_by(Province_State,Country_Region,date) %>%
  summarize(cases=sum(cases), deaths=sum(deaths), Population= sum(Population)) %>%
  mutate(deaths_per_mill = deaths*1000000/Population) %>%
  select(Province_State,Country_Region,date,cases,deaths,deaths_per_mill,Population) %>% ungroup()

US_totals <- US_by_state %>% 
  group_by(Country_Region,date) %>% summarize(cases=sum(cases),deaths=sum(deaths),Population = sum(Population)) %>% mutate(deaths_per_mill = deaths*1000000/Population) %>% select(Country_Region,date,cases,deaths,deaths_per_mill,Population) %>%ungroup()

US_totals %>% filter(cases >0) %>% 
  ggplot(aes(x=date,y=cases)) + geom_line(aes(color = "cases")) + geom_point(aes(color = "cases")) + geom_line(aes(y=deaths,color="deaths")) + geom_point(aes(y=deaths,color="deaths")) + scale_y_log10() +
  theme(legend.position="bottom",axis.text.x = element_text(angle=90)) +
  labs(title="COVID19 in US", y= NULL)

```


```{r us_by_state, echo=TRUE}
#Filter for case in New York
state <- "New York" 
US_by_state %>%
  filter(Province_State == state ) %>% filter(cases>0) %>%
  ggplot(aes(x=date,y=cases)) +
  geom_line(aes(color="cases")) + geom_point(aes(color="cases")) +
  geom_line(aes(y=deaths,color="deaths")) + 
  geom_point(aes(y=deaths, color="deaths")) +
  geom_point(aes(y=deaths,color="deaths")) +
  scale_y_log10() +
  theme(legend.position="bottom", axis.text.x = element_text(angle=90)) +
  labs(title=str_c("COVID19 in ", state),y=NULL )

```

```{r us_by_state2, echo=TRUE}
US_by_state <- US_by_state %>% 
  mutate(new_cases = cases - lag(cases),
         new_deaths = deaths-lag(deaths))

US_totals <- US_totals %>% 
  mutate(new_cases = cases - lag(cases),new_deaths=deaths-lag(deaths))
```

```{r us_totals2, echo=TRUE}
US_totals %>% 
  ggplot(aes(x=date,y=new_cases)) + geom_line(aes(color = "new_cases")) + geom_point(aes(color = "new_cases")) + geom_line(aes(y=new_deaths,color="new_deaths")) + geom_point(aes(y=new_deaths,color="new_deaths")) + scale_y_log10() +
  theme(legend.position="bottom",axis.text.x = element_text(angle=90)) +
  labs(title="COVID19 in US", y= NULL)

```

```{r us_by_state3, echo=TRUE}
state <- "Texas" 

US_by_state %>%
  filter(Province_State == state ) %>% 
  ggplot(aes(x=date,y=new_cases)) +
  geom_line(aes(color="new_cases")) + geom_point(aes(color="new_cases")) +
  geom_line(aes(y=new_deaths,color="new_deaths")) + 
  geom_point(aes(y=new_deaths, color="new_deaths")) +
  geom_point(aes(y=new_deaths,color="new_deaths")) +
  scale_y_log10() +
  theme(legend.position="bottom", axis.text.x = element_text(angle=90)) +
  labs(title=str_c("COVID19 in ", state),y=NULL )

```

```{r us_state_totals, echo=TRUE}
US_state_totals<- US_by_state %>% 
  group_by(Province_State) %>% 
  summarize(deaths = max(deaths),cases = max(cases), population= max(Population),
            cases_per_thou = 1000*cases/population,
            deaths_per_thou=1000*deaths/population) %>%
  filter(cases >0,population >0)
US_state_totals %>% slice_min(deaths_per_thou,n=10)

US_state_totals %>% slice_min(deaths_per_thou,n=10) %>%
  select(deaths_per_thou,cases_per_thou,everything())

US_state_totals %>% slice_max(deaths_per_thou,n=10) %>%
  select(deaths_per_thou,cases_per_thou,everything())

```
### Cases in France and it's territories

```{r france, echo=TRUE}
france <- global %>% 
  filter(Country_Region =="France")
france
summary(france)
france_by_state <- france %>% group_by(Province_State,Country_Region,date) %>%
  summarize(cases=sum(cases), deaths=sum(deaths), Population= sum(Population)) %>%
  mutate(deaths_per_mill = deaths*1000000/Population) %>%
  select(Province_State,Country_Region,date,cases,deaths,deaths_per_mill,Population) %>% ungroup()
france_by_state

```

```{r france_totals, echo=TRUE}
france_totals <- france_by_state %>% group_by(Country_Region,date) %>% summarize(cases=sum(cases),deaths=sum(deaths),Population = sum(Population)) %>% mutate(deaths_per_mill = deaths*1000000/Population) %>% select(Country_Region,date,cases,deaths,deaths_per_mill,Population) %>%ungroup()
```

```{r france_totals2, echo=TRUE}
france_totals %>% filter(cases >0) %>% 
  ggplot(aes(x=date,y=cases)) + geom_line(aes(color = "cases")) + geom_point(aes(color = "cases")) + geom_line(aes(y=deaths,color="deaths")) + geom_point(aes(y=deaths,color="deaths")) + scale_y_log10() +
  theme(legend.position="bottom",axis.text.x = element_text(angle=90)) +
  labs(title="COVID19 in France", y= NULL)

```

```{r france_state_totals3, echo=TRUE}
france_state_totals<- france_by_state %>% 
  group_by(Province_State) %>% 
  summarize(deaths = max(deaths),cases = max(cases), population= max(Population),
            cases_per_thou = 1000*cases/population,
            deaths_per_thou=1000*deaths/population) %>%
  filter(cases >0,population >0)
mod_fran <- lm(deaths_per_thou ~cases_per_thou,data=france_state_totals)
france_state_totals %>% mutate(pred=predict(mod_fran))
france_state_totals <- france_state_totals %>% replace_na(list(Province_State ="France"))
france_state_totals
france_tot_w_pred <- france_state_totals %>% mutate(pred=predict(mod_fran))
france_tot_w_pred %>% ggplot() + 
  geom_point(aes(x=cases_per_thou,y=deaths_per_thou),color="blue") +
  geom_point(aes(x=cases_per_thou,y=pred), color="red")

```

```{r france_totals4, echo=TRUE}
france_totals <- france_totals %>% 
  mutate(new_cases = cases - lag(cases),new_deaths=deaths-lag(deaths))

france_totals %>% 
  ggplot(aes(x=date,y=new_cases)) + geom_line(aes(color = "new_cases")) + geom_point(aes(color = "new_cases")) + geom_line(aes(y=new_deaths,color="new_deaths")) + geom_point(aes(y=new_deaths,color="new_deaths")) + scale_y_log10() +
  theme(legend.position="bottom",axis.text.x = element_text(angle=90)) +
  labs(title="COVID19 in France", y= NULL)

```

The analysis reveals that the number of new cases and deaths in France and its territories are still increasing.

## Conclusion
Based on the analysis of COVID19 data for France and its overseas territories, it is evident that there is a rise in the number of new cases and deaths. This indicates an ongoing spread and impact of the virus in the region. It is crucial to continue monitoring and implementing necessary measures to control and mitigate the further transmission of COVID19.
It is important to note that this analysis is based on the available data from the John Hopkins University GitHub account. Any biases in the data collection or reporting methods could potentially affect the accuracy and reliability of the findings. Additionally, the analysis focuses specifically on France and may not capture the complete global picture of the COVID19 situation.In mitigating the my biases on the data, the analysis and report was void of any personal assumptions and interpretation of reasons for the increased number of cases and deaths. 
